### 2010-04-01 papers/icfp-2010-desc/paper_desc_levitation.tex
Desc@ICFP: WeightWatch section 4, Levitation

In this section, we will fulfil several promises. One promise we made
was to effectively implement the code of the $\Desc$ universe. Another
promise we tacitly made was the existence of the type formers for
finite sets. Despite being a perilous pedagogical exercise for the
authors, we hope to convey to the reader the dizzy feeling of
levitation, without the falling.

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. It consists of the objects presented in
Figure~\ref{fig:typing-finite-set}. We are going to implement the
$\EnumU$ type former and its constructors. Let us recall their
specification:

@@

The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
is indeed isomorphic to a list of $\UId$s. In
Section~\ref{sec:desc-examples}, we have shown how to build lists in
the universe of description. Therefore, the actual implementation of
$\EnumU$ boils down to the following, trivial definition:

@@

Let us examine the consequences. First of all, we discover that the
type theory does not actually need to be extended with the type former
$\EnumU$, the constructors $\NilE$ and $\ConsE$. We have just defined
$\EnumU$ above, by fix-point over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that we
remove an operator hard-coded in the host language. This operator is
simply implemented in the dependently-typed target language, as any
other function.

Apart from these changes, we are left with implementing the other
components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
well as the $\F{switch}$ operator. Note that the actual implementation
of $\EnumU$ does not influence our implementation of $\EnumT$: be they
hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
behave the same. We have witnessed this effect when carrying the
operation in Epigram, moving from an hard-coded presentation to a
self-hosted one. Absolutely no change to the $\EnumT$ objects was
required.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the datatypes we have seen so far, we first have to choose a
constructor. Constructors correspond to the code's constructors:
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are
defining.

The case of $\DUnit$ is trivial: we immediately reach the end of the
datatype. $\DUnit$ has no payload. The case of $\DIndx{\!}$ comes
naturally: it only takes a recursive argument -- a description. The
case of $\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication and do not replace $\DIndx{\!}$ by the
more general $\DHindx{\!}{\!}$. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Therefore, while both codes are
\emph{extensionally} the same, in most practical implementations they
would be dealt with rather differently. While definitional equality
can cope with first order objects, the functional presentation
introduced by $\DHindx{\!}{}$ is unlikely to be amenable to a purely
definitional treatment.

Equipped with $\DHindx{\!}{\!}$, we can describe the extended code of
our universe of datatypes, as shown in
Figure~\ref{fig:desc-levitate}.  The $\DUnit$ and $\DIndx{\!}$ cases
remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
definition, as formalised in Figure~\ref{fig:type-checking}.

@@

If we cannot make it happen, we simply do not make it happen. We
perfectly know what the motive ought to be. Consequently, we extend
the type theory with a special-purpose operator:

@@

The entire work of the magician stands here, in this extension. One
could be worried by the availability of such operations in our type
theory. Indeed, it would be inconvenient for the developer to have to
decide between two variants. The solution to this problem is, once
again, to use the type propagation system. Just as the $\F{switch}$
operator, $\F{switchD}$ can be automatically elaborated. To this end,
we extend the type-checking rules (Fig.~\ref{fig:type-checking}) with
the following inference rule:

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, this exercise
reveals that the $\Desc$ universe is just plain data. Just as any
piece of data, it can therefore be inspected and
manipulated. Moreover, it is expressed in the $\Desc$ universe. As a
consequence, it is equipped, for free, with an induction
principle. So, our ability to inspect and program with $\Desc$ is not
restricted to the meta-language: we now have all the necessary
equipment in the target language to \emph{program} over datatypes. In
this setting, \emph{generic programming is just about programming}.

@@

In Section~\ref{sec:desc-fix-point}, we have argued for the
implementation of a dependently-typed $\F{induction}$ principle,
instead of the more traditional catamorphism. However, in some
circumstances, the full-power of a dependent elimination is not
necessary. In the following, we propose to derive the catamorphism
from the generic induction principle.

The catamorphism can be characterised as an induction on the
description $\V{D}$, with a non-dependent motive targeting $\V{T}$. Given a
node $\V{xs}$ and the induction hypotheses, the method ought to build an
element of $\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra $\V{f}$. Let
us take a look at this jigsaw:

@@

We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$. Hence the name ``induction
hypotheses''. Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
\(\V{xs}\) by their counterpart in \(\V{hs}\):

@@

The astute reader will have been struck by the type of $\F{replace}$:
it is \emph{almost} the morphism part -- sometimes called \emph{map}
-- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
is the dependent version of the map, which uses the induction
hypotheses. For space reason, we will not present the non-dependent
map. It can be found in the Agda model.

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a datatype specific
operation. This programming style is reminiscent to polytypic
languages, such as PolyP~\cite{jansson:polyp}.

Moreover, the $\F{replace}$ function demonstrates the benefit of an
approach based on universes. The datatypes living in the universe of
descriptions, we are able to \emph{inspect} them. As shown by
$\F{replace}$, it is easy to explore these structures, as well as
building new ones.

@@

\note{Recall monadic structure of this object}

@@

In our setting, the functor will be represented by a tagged
description. The free monad construction will take such tagged
description, a set $\V{X}$ of variables, and will compute the tagged
description of the corresponding free monad. Implementing this
function is surprisingly easy:

@@

Our task is therefore to implement the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. Whereas the \return\ is a trivial variable former, the
\bind\ operation corresponds to the substitution of variables by
terms. We have shown how to implement it, for any free monad. This
programming style is reminiscent to type-indexed
datatypes~\cite{hinze:generic-haskell}, as implemented in Generic
Haskell: from datatypes, we built new datatypes and equip them with
their structure.

@@

%% Candidate for removal:
With the free monad construction, we have seen two kinds of generic
operations. Firstly, we have derived a new data-structure from another
one: we make the free monad from its underlying functor. To do so, we
crucially rely on the fact that datatypes are nothing but codes. We
are therefore entitled to modify this code and, in this case, extend
it. Extending a datatype might give rise to a more structured object,
as was the case here.  So, secondly, we have equipped this new
datatype with its inherent structure: the \bind\ and
\return\ operations. We have been able to build them as generic
functions.

>>>>

In this section, we will fulfil our promises. First, we will build
finite sets. Second, we will implement the code of the $\Desc$
universe.  Despite being a perilous pedagogical exercise for the
authors, we hope to convey to the reader the dizzy feeling of
levitation, without the falling.

@@

In Section~\ref{sec:finite-sets}, we have given a specification of
finite sets. We are going to implement the $\EnumU$ type former and
its constructors. Let us recall their specification:

@@

The names $\NilE$ and $\ConsE{\!}{\!}$ are not a coincidence. $\EnumU$
is indeed isomorphic to a list of $\UId$s. Therefore, the actual
implementation of $\EnumU$ boils down to the following definition:

@@

Let us consider the consequences. First, we discover that the type
theory does not need to be extended with the type former $\EnumU$, the
constructors $\NilE$ and $\ConsE$. We have just defined $\EnumU$
above, by fix-point over the pattern functor denoted by
$\ListD$. $\NilE$ and $\ConsE$ are simply the $\ListNil$ and
$\ListCons$ constructors of lists.

Another interesting consequence concerns the $\spi{\!}{\!}$ operator:
it can actually be implemented with the generic $\F{induction}$
principle. From an implementation perspective, this means that this
operator is simply implemented in the dependently-typed target
language, as any other function.

%% Apart from these changes, we are left with implementing the other
%% components of finite sets. This consists of $\EnumT$, $\Ze$, $\Su$, as
%% well as the $\F{switch}$ operator. Note that the actual implementation
%% of $\EnumU$ does not influence our implementation of $\EnumT$: be they
%% hard-coded or codes in the $\Desc$ universe, the $\EnumU$ objects
%% behave the same. We have witnessed this effect when carrying the
%% operation in Epigram, moving from an hard-coded presentation to a
%% self-hosted one. Absolutely no change to the $\EnumT$ objects was
%% required.

@@

Let us explain how we have proceeded and the obstacle we face. Much as
the datatypes we have seen so far, we first choose a constructor among
$\DUnit$, $\DSigma{\!}{\!}$, and $\DIndx{\!}$. The reader will notice
that the tagged notation we have used for the $\Desc$ constructors now
fully makes sense: these were actually the tags we are defining.

For $\DUnit$, we immediately reach the end of the description. For
$\DIndx{\!}$, there is a single recursive argument. The case of
$\DSigma{\!}{\!}$ is problematic. Recall the specification of
$\DSigma{\!}{\!}$:

@@

Note that $\DIndx{\!}$ is isomorphic to $\DHindx{\Unit}{\!}$. However,
we tolerate this duplication. Indeed, unlike its counterpart,
$\DIndx{\!}$ is first-order. Both codes are \emph{extensionally} the
same but, in practice, they will be dealt with rather
differently. While definitional equality can cope with first order
objects, the functional presentation introduced by $\DHindx{\!}{}$ is
unlikely to be amenable to a purely definitional treatment.

As shown in Figure~\ref{fig:desc-levitate}, we are now able to
describe our universe of datatypes.  The $\DUnit$ and $\DIndx{\!}$
cases remain unchanged, as expected. We successfully describe the
$\DSigma{\!}{}$ case, by a simple appeal to the higher-order induction
on $\V{S}$. The $\DHindx{\!}{\!}$ case consists in packing the $\V{H}$
in $\Set$ with a recursive argument.

@@

Our reader might be slightly confused to learn that the trick is
visible in the definition of $\DescD$
(Fig.~\ref{fig:desc-levitate}). Or rather, it is made invisible by
careful usage of type propagation. Indeed, let us try to elaborate
this term down to the low-level type theory. The $\EnumT{[ \ldots ]}$
construct elaborates to a finite set $e$ in $\EnumU$, inhabited by the
codes. Then, we can type-check the case definition, between square
brackets $[ \ldots ]$. This term $\V{f}$ is pushed into the type
$\EnumT{\V{e}} \To \Mu{\DescD}$. This corresponds to a finite function
definition, as presented in Figure~\ref{fig:type-checking}.

@@

However, we perfectly know what the motive is. Consequently, we extend
the type theory with a special-purpose operator:

@@

The entire work of the magician stands here, in this extension. Note
that the developer do not have to decide between the two variants of
$\F{switch}$. This is dealt with by the type propagation system
(Fig.~\ref{fig:type-checking}), extended by the following inference
rule:

@@

This concludes our levitation work. Beyond its pedagogical value, this
exercise has several practical outcomes. First of all, it reveals that
the $\Desc$ universe is just plain data. Just as any piece of data, it
can therefore be inspected and manipulated. Moreover, it is expressed
in the $\Desc$ universe. As a consequence, it is equipped, for free,
with an induction principle. So, our ability to inspect and program
with $\Desc$ is not restricted to the meta-language: we now have all
the necessary equipment in the target language to \emph{program} over
datatypes. In this setting, \emph{generic programming is just 
  programming}.

@@

In Section~\ref{sec:desc-fix-point}, we have implemented a dependent
$\F{induction}$ principle, instead of the catamorphism. However, in
some circumstances, the full-power of a dependent elimination is not
necessary. In the following, we propose to derive the catamorphism
from the $\F{induction}$ principle.

The catamorphism is an induction on the description $\V{D}$, with a
non-dependent motive targeting $\V{T}$. Given a node $\V{xs}$ and the
induction hypotheses, the method ought to build an element of
$\V{T}$. Provided that we know how to make an element of
$\descop{\V{D}}{\V{T}}$, this step will be performed by the algebra
$\V{f}$. Let us take a look at this jigsaw:

@@

We are left with filling the hole. Recall that we have
\(\Bhab{\V{xs}}{\descop{\V{D}}{\Mu{\V{D}}}}\) and
\(\Bhab{\V{hs}}{\All{\V{D}}{(\Mu{\V{D}})}{(\LAM{\_} \V{T})}{\V{xs}}}\)
at hand. Our goal is to make an element of
\(\descop{\V{D}}{\V{T}}\). Intuitively, $\V{xs}$ is of the right
shape, but its sub-elements are of the wrong type. On the other hand,
for each sub-element of $\V{xs}$, $\V{hs}$ gives us the corresponding
element in $\V{T}$.  Therefore, to construct an element of
\(\descop{\V{D}}{\V{T}}\), we replace the recursive components of
\(\V{xs}\) by their counterpart in \(\V{hs}\):

@@

%% The astute reader will have been struck by the type of $\F{replace}$:
%% it is \emph{almost} the morphism part -- sometimes called \emph{map}
%% -- of the functor $\V{D}$ from $\V{X}$ to $\V{Y}$ in $\Set$. Just as
%% the $\F{induction}$ is the dependent version of $\cata$, $\F{replace}$
%% is the dependent version of the map, which uses the induction
%% hypotheses. For space reason, we will not present the non-dependent
%% map. It can be found in the Agda model.

@@

With this example, we have shown how we can derive a generic
operation, the catamorphism, from a pre-existing generic operation,
the induction principle. This has been made possible by our ability to
manipulate descriptions as first-class objects: the catamorphism is,
basically, a function mapping a $\Desc$ to a datatype specific
operation. This is a form of polytypic programming, as found in
PolyP~\cite{jansson:polyp}.

%% Moreover, the $\F{replace}$ function demonstrates the benefit of an
%% approach based on universes. The datatypes living in the universe of
%% descriptions, we are able to \emph{inspect} them. As shown by
%% $\F{replace}$, it is easy to explore these structures, as well as
%% building new ones.

@@

@@

In our setting, the free monad construction will take the functor as a
tagged description, a set $\V{X}$ of variables, and will compute the
tagged description of the corresponding free monad. Implementing this
function is surprisingly easy:

@@

We are left with implementing the algebra of the
catamorphism. Intuitively, its role is to catch appearances of
$\DVar{\V{x}}$ and replace them by $\V{\sigma}\: \V{x}$. This
corresponds to the following definition:

@@

Filling the sub-goal with $\apply\: \V{D}\: \V{X}\: \V{Y}\:
\V{\sigma}$ completes the implementation. To sum up, we have
implemented the free monad construction for an arbitrary tagged
description. This gives the developer the ability, for any datatype,
to extend it with a notion of variable. Then, we have equipped this
structure with the corresponding monadic operation, \bind\ and
\return. We have shown how to implement \bind, for any free
monad. This construction is an example of type-indexed
datatype~\cite{hinze:generic-haskell}, as found in Generic Haskell:
from a datatype, we built a new datatype and equip it with its
structure.

@@

%% Candidate for removal:
%% With the free monad construction, we have seen two kinds of generic
%% operations. Firstly, we have derived a new data-structure from another
%% one: we make the free monad from its underlying functor. To do so, we
%% crucially rely on the fact that datatypes are nothing but codes. We
%% are therefore entitled to modify this code and, in this case, extend
%% it. Extending a datatype might give rise to a more structured object,
%% as was the case here.  So, secondly, we have equipped this new
%% datatype with its inherent structure: the \bind\ and
%% \return\ operations. We have been able to build them as generic
%% functions.
